---
title: "P8106 Midterm Project – EDA & Data Cleaning"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.path   = "figures/"
)

library(MASS)
library(corrplot)
library(caret)
library(glmnet)
library(tidymodels)
```

# 1) Data Loading

```{r}
load("data/datA.RData")
load("data/datB.RData")

dim(datA)
dim(datB)
```

# 2) Data Cleaning

## 2.1 Factor Encoding

Encode all categorical variables as factors with interpretable labels.

```{r}
convert_factors <- function(df) {
  df$menopause       <- factor(df$menopause, levels = 0:1,
                               labels = c("Pre", "Post"))
  df$race            <- factor(df$race, levels = 1:4,
                               labels = c("White", "Black", "Asian", "Hispanic"))
  df$tumor_grade     <- factor(df$tumor_grade, levels = 1:3,
                               labels = c("Well", "Moderate", "Poor"))
  df$stage           <- factor(df$stage, levels = 1:3,
                               labels = c("I", "II", "III"))
  df$ER              <- factor(df$ER, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$PR              <- factor(df$PR, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$HER2            <- factor(df$HER2, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$surgery         <- factor(df$surgery, levels = 0:1,
                               labels = c("BCS", "Mastectomy"))
  df$chemo           <- factor(df$chemo, levels = 0:1,
                               labels = c("No", "Yes"))
  df$hormonal_therapy <- factor(df$hormonal_therapy, levels = 0:1,
                                labels = c("No", "Yes"))
  return(df)
}

datA <- convert_factors(datA)
datB <- convert_factors(datB)

str(datA)
```

## 2.2 Missing Data Check

```{r}
colSums(is.na(datA))
colSums(is.na(datB))
```

No missing values in either dataset — no imputation needed.

# 3) Exploratory Data Analysis (Cohort A)

## 3.1 Summary Statistics

```{r}
summary(datA[, -1])
```

## 3.2 Response Variable Distribution

```{r}
par(mfrow = c(1, 2))

hist(datA$log_PFS, breaks = 30,
     main = "Histogram of log(PFS)",
     xlab = "log(PFS)", col = "steelblue", border = "white")
abline(v = mean(datA$log_PFS), col = "red", lwd = 2, lty = 2)
legend("topright",
       legend = paste("Mean =", round(mean(datA$log_PFS), 3)),
       lty = 2, col = "red", lwd = 2, bty = "n")

qqnorm(datA$log_PFS, main = "Q-Q Plot of log(PFS)")
qqline(datA$log_PFS, col = "red", lwd = 2)
```

log(PFS) is approximately normal with mean `r round(mean(datA$log_PFS), 3)` and SD `r round(sd(datA$log_PFS), 3)`.

## 3.3 Continuous Predictors

```{r, fig.height=4}
cont_vars <- c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")

par(mfrow = c(1, 5), mar = c(4, 3, 2, 1))
for (v in cont_vars) {
  boxplot(datA[[v]], main = v, col = "steelblue", outline = TRUE)
}
```

```{r}
cont_df <- datA[, cont_vars]
data.frame(
  Variable = cont_vars,
  Mean     = round(sapply(cont_df, mean), 2),
  SD       = round(sapply(cont_df, sd), 2),
  Median   = round(sapply(cont_df, median), 2),
  Min      = round(sapply(cont_df, min), 2),
  Max      = round(sapply(cont_df, max), 2)
)
```

## 3.4 Correlation Matrix

```{r, fig.height=5}
cont_with_y <- c(cont_vars, "log_PFS")
cor_mat <- cor(datA[, cont_with_y])

corrplot(cor_mat,
         method  = "color",
         type    = "lower",
         addCoef.col = "black",
         number.cex  = 0.8,
         tl.col  = "black",
         tl.cex  = 0.9,
         title   = "Correlation Matrix (Continuous Variables)",
         mar     = c(0, 0, 2, 0))
```

Key correlations with `log_PFS`:

- **tumor_size** (r = `r round(cor_mat["tumor_size", "log_PFS"], 3)`) — moderate negative
- **lymph_nodes** (r = `r round(cor_mat["lymph_nodes", "log_PFS"], 3)`) — moderate negative
- **Ki67** (r = `r round(cor_mat["Ki67", "log_PFS"], 3)`) — moderate negative

Note: `tumor_size` and `lymph_nodes` are strongly correlated with each other (r = `r round(cor_mat["tumor_size", "lymph_nodes"], 3)`), suggesting potential multicollinearity.

## 3.5 Categorical Variable Distributions

```{r, fig.height=8}
cat_vars <- c("menopause", "race", "tumor_grade", "stage",
              "ER", "PR", "HER2",
              "surgery", "chemo", "hormonal_therapy")

par(mfrow = c(3, 4), mar = c(4, 3, 2, 1))
for (v in cat_vars) {
  barplot(table(datA[[v]]),
          main = v, col = "steelblue", border = "white",
          las = 2, cex.names = 0.8)
}
```

```{r}
lapply(cat_vars, function(v) {
  tbl <- table(datA[[v]])
  pct <- round(prop.table(tbl) * 100, 1)
  data.frame(Variable = v, Level = names(tbl), N = as.integer(tbl), Pct = paste0(pct, "%"))
}) |> do.call(rbind, args = _)
```

## 3.6 Outcome by Key Predictors

```{r, fig.height=8}
key_cats <- c("ER", "PR", "HER2", "tumor_grade", "stage", "hormonal_therapy")

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (v in key_cats) {
  boxplot(log_PFS ~ datA[[v]], data = datA,
          main = paste("log(PFS) by", v),
          xlab = v, ylab = "log(PFS)",
          col = "steelblue", border = "gray30")
}
```

## 3.7 Scatterplots: Continuous Predictors vs log(PFS)

```{r, fig.height=4}
par(mfrow = c(1, 5), mar = c(4, 4, 2, 1))
for (v in cont_vars) {
  plot(datA[[v]], datA$log_PFS,
       pch = 16, cex = 0.4, col = rgb(0, 0.3, 0.6, 0.3),
       xlab = v, ylab = "log(PFS)",
       main = paste(v, "vs log(PFS)"))
  abline(lm(datA$log_PFS ~ datA[[v]]), col = "red", lwd = 2)
}
```

---

*Next steps: model training, diagnostics, and external validation.*

## Model Training 
```{r}
# creating train/test split
set.seed(1)
n = nrow(datA)
data_split = initial_split(datA, prop = 0.8)
trainA = training(data_split)
trainA = subset(trainA, select = -id)
testA = testing(data_split)
testA = subset(testA, select = -id)
ctrl = trainControl(method = "cv", number = 10)
```

### Fitting the Linear Regression
```{r}
set.seed(1)
fit_lm = train(log_PFS ~ ., data = trainA, method = "lm", trControl = ctrl)
fit_lm
summary(fit_lm$finalModel)
pred_lm = predict(fit_lm, newdata = testA)
lm_result = postResample(pred = pred_lm, obs = testA$log_PFS)
```

### Fitting the glmnet model
```{r}
set.seed(1)
fit_glmnet = train(log_PFS ~ ., data = trainA, method = "glmnet", trControl = ctrl, tuneLength = 10)
fit_glmnet
plot(fit_glmnet)
fit_glmnet$bestTune
pred_glmnet = predict(fit_glmnet, newdata = testA)
glmnet_result = postResample(pred = pred_glmnet, obs = testA$log_PFS)
```
### Comparing the two models 
```{r}

results <- data.frame(lm_result, glmnet_result)
results
coef(fit_glmnet$finalModel, fit_glmnet$bestTune$lambda)
```

### Result
fitting model with `glmnet` framework shows a better performance because it has lower RMSE & MAE, and a higher R-squared. We selected the elastic-net model as the final model for Cohort A. t also produced a more parsimonious model by shrinking some coefficients to zero, including BMI, PR, stage II, and chemotherapy. The predictors that remained important in the final model were age, menopausal status, race, tumor size, tumor grade, lymph nodes, ER, HER2, Ki67, surgery type, and hormonal therapy.

### Interpretation and Assessment
Overall, the final model suggests that greater tumor burden and more aggressive tumor biology are associated with shorter progression-free survival, while ER positivity and hormonal therapy are associated with longer progression-free survival. However, predictive performance was only modest (test-set R² about 0.17), which means the available baseline variables explain some, but not all, of the variation in log_PFS.

So for the next part, the elastic-net model from Cohort A should be carried forward for external validation on Cohort B.



