---
title: "P8106 Midterm Project – EDA & Data Cleaning"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.path   = "figures/"
)

library(MASS)
library(corrplot)
```

# 1 Data Loading

```{r load-data}
load("data/datA.RData")
load("data/datB.RData")

cat("datA:", nrow(datA), "obs x", ncol(datA), "vars\n")
cat("datB:", nrow(datB), "obs x", ncol(datB), "vars\n")
```

# 2 Data Cleaning

## 2.1 Factor Encoding

Encode all categorical variables as factors with interpretable labels.

```{r factor-encoding}
convert_factors <- function(df) {
  df$menopause       <- factor(df$menopause, levels = 0:1,
                               labels = c("Pre", "Post"))
  df$race            <- factor(df$race, levels = 1:4,
                               labels = c("White", "Black", "Asian", "Hispanic"))
  df$tumor_grade     <- factor(df$tumor_grade, levels = 1:3,
                               labels = c("Well", "Moderate", "Poor"))
  df$stage           <- factor(df$stage, levels = 1:3,
                               labels = c("I", "II", "III"))
  df$ER              <- factor(df$ER, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$PR              <- factor(df$PR, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$HER2            <- factor(df$HER2, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$surgery         <- factor(df$surgery, levels = 0:1,
                               labels = c("BCS", "Mastectomy"))
  df$chemo           <- factor(df$chemo, levels = 0:1,
                               labels = c("No", "Yes"))
  df$hormonal_therapy <- factor(df$hormonal_therapy, levels = 0:1,
                                labels = c("No", "Yes"))
  return(df)
}

datA <- convert_factors(datA)
datB <- convert_factors(datB)

str(datA)
```

## 2.2 Missing Data Check

```{r missing-data}
cat("--- datA ---\n")
colSums(is.na(datA))

cat("\n--- datB ---\n")
colSums(is.na(datB))
```

No missing values in either dataset — no imputation needed.

# 3 Exploratory Data Analysis (Cohort A)

## 3.1 Summary Statistics

```{r summary-stats}
summary(datA[, -1])
```

## 3.2 Response Variable Distribution

```{r response-dist, fig.cap="Distribution of log(PFS) in Cohort A"}
par(mfrow = c(1, 2))

hist(datA$log_PFS, breaks = 30,
     main = "Histogram of log(PFS)",
     xlab = "log(PFS)", col = "steelblue", border = "white")
abline(v = mean(datA$log_PFS), col = "red", lwd = 2, lty = 2)
legend("topright",
       legend = paste("Mean =", round(mean(datA$log_PFS), 3)),
       lty = 2, col = "red", lwd = 2, bty = "n")

qqnorm(datA$log_PFS, main = "Q-Q Plot of log(PFS)")
qqline(datA$log_PFS, col = "red", lwd = 2)
```

log(PFS) is approximately normal with mean `r round(mean(datA$log_PFS), 3)` and SD `r round(sd(datA$log_PFS), 3)`.

## 3.3 Continuous Predictors

```{r continuous-boxplots, fig.height=4}
cont_vars <- c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")

par(mfrow = c(1, 5), mar = c(4, 3, 2, 1))
for (v in cont_vars) {
  boxplot(datA[[v]], main = v, col = "steelblue", outline = TRUE)
}
```

```{r continuous-summary}
cont_df <- datA[, cont_vars]
data.frame(
  Variable = cont_vars,
  Mean     = round(sapply(cont_df, mean), 2),
  SD       = round(sapply(cont_df, sd), 2),
  Median   = round(sapply(cont_df, median), 2),
  Min      = round(sapply(cont_df, min), 2),
  Max      = round(sapply(cont_df, max), 2)
)
```

## 3.4 Correlation Matrix

```{r correlation-matrix, fig.height=5}
cont_with_y <- c(cont_vars, "log_PFS")
cor_mat <- cor(datA[, cont_with_y])

corrplot(cor_mat,
         method  = "color",
         type    = "lower",
         addCoef.col = "black",
         number.cex  = 0.8,
         tl.col  = "black",
         tl.cex  = 0.9,
         title   = "Correlation Matrix (Continuous Variables)",
         mar     = c(0, 0, 2, 0))
```

Key correlations with `log_PFS`:

- **tumor_size** (r = `r round(cor_mat["tumor_size", "log_PFS"], 3)`) — moderate negative
- **lymph_nodes** (r = `r round(cor_mat["lymph_nodes", "log_PFS"], 3)`) — moderate negative
- **Ki67** (r = `r round(cor_mat["Ki67", "log_PFS"], 3)`) — moderate negative

Note: `tumor_size` and `lymph_nodes` are strongly correlated with each other (r = `r round(cor_mat["tumor_size", "lymph_nodes"], 3)`), suggesting potential multicollinearity.

## 3.5 Categorical Variable Distributions

```{r categorical-barplots, fig.height=8}
cat_vars <- c("menopause", "race", "tumor_grade", "stage",
              "ER", "PR", "HER2",
              "surgery", "chemo", "hormonal_therapy")

par(mfrow = c(3, 4), mar = c(4, 3, 2, 1))
for (v in cat_vars) {
  barplot(table(datA[[v]]),
          main = v, col = "steelblue", border = "white",
          las = 2, cex.names = 0.8)
}
```

```{r categorical-tables}
for (v in cat_vars) {
  cat(sprintf("\n--- %s ---\n", v))
  tbl <- table(datA[[v]])
  pct <- round(prop.table(tbl) * 100, 1)
  print(data.frame(Level = names(tbl), N = as.integer(tbl), Pct = paste0(pct, "%")))
}
```

## 3.6 Outcome by Key Predictors

```{r boxplots-by-category, fig.height=8}
key_cats <- c("ER", "PR", "HER2", "tumor_grade", "stage", "hormonal_therapy")

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (v in key_cats) {
  boxplot(log_PFS ~ datA[[v]], data = datA,
          main = paste("log(PFS) by", v),
          xlab = v, ylab = "log(PFS)",
          col = "steelblue", border = "gray30")
}
```

## 3.7 Scatterplots: Continuous Predictors vs log(PFS)

```{r scatter-vs-pfs, fig.height=4}
par(mfrow = c(1, 5), mar = c(4, 4, 2, 1))
for (v in cont_vars) {
  plot(datA[[v]], datA$log_PFS,
       pch = 16, cex = 0.4, col = rgb(0, 0.3, 0.6, 0.3),
       xlab = v, ylab = "log(PFS)",
       main = paste(v, "vs log(PFS)"))
  abline(lm(datA$log_PFS ~ datA[[v]]), col = "red", lwd = 2)
}
```

# 4 Cohort Comparison (A vs B)

Quick comparison to flag distributional shifts between the 1990–1995 and 2000–2005 cohorts.

## 4.1 Outcome

```{r cohort-outcome, fig.height=4}
par(mfrow = c(1, 2))
hist(datA$log_PFS, breaks = 30, xlim = c(-1, 6),
     main = "Cohort A (1990-1995)",
     xlab = "log(PFS)", col = "steelblue", border = "white")
abline(v = mean(datA$log_PFS), col = "red", lwd = 2, lty = 2)

hist(datB$log_PFS, breaks = 30, xlim = c(-1, 6),
     main = "Cohort B (2000-2005)",
     xlab = "log(PFS)", col = "coral", border = "white")
abline(v = mean(datB$log_PFS), col = "red", lwd = 2, lty = 2)
```

```{r cohort-outcome-test}
t.test(datA$log_PFS, datB$log_PFS)
```

## 4.2 Continuous Predictors

```{r cohort-continuous}
for (v in cont_vars) {
  tt <- t.test(datA[[v]], datB[[v]])
  cat(sprintf("%-12s  A: %.2f (%.2f)  B: %.2f (%.2f)  p = %.4f\n",
              v,
              mean(datA[[v]]), sd(datA[[v]]),
              mean(datB[[v]]), sd(datB[[v]]),
              tt$p.value))
}
```

## 4.3 Categorical Predictors

```{r cohort-categorical}
for (v in cat_vars) {
  tab <- rbind(A = table(datA[[v]]), B = table(datB[[v]]))
  chi <- chisq.test(tab)
  cat(sprintf("%-18s  chi-sq = %6.2f  p = %.4f\n", v, chi$statistic, chi$p.value))
}
```

---

*Next steps: model training, diagnostics, and external validation will be added in a separate analysis file.*
