---
title: "P8106 Midterm Project – EDA, Modeling & Generalization"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.path   = "figures/"
)

library(MASS)
library(corrplot)
library(caret)
library(glmnet)
library(tidymodels)
library(mgcv)
```

# 1 Data Loading

```{r}
load("data/datA.RData")
load("data/datB.RData")

dim(datA)
dim(datB)
```

# 2 Data Cleaning

## 2.1 Factor Encoding

Encode all categorical variables as factors with interpretable labels.

```{r}
convert_factors <- function(df) {
  df$menopause       <- factor(df$menopause, levels = 0:1,
                               labels = c("Pre", "Post"))
  df$race            <- factor(df$race, levels = 1:4,
                               labels = c("White", "Black", "Asian", "Hispanic"))
  df$tumor_grade     <- factor(df$tumor_grade, levels = 1:3,
                               labels = c("Well", "Moderate", "Poor"))
  df$stage           <- factor(df$stage, levels = 1:3,
                               labels = c("I", "II", "III"))
  df$ER              <- factor(df$ER, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$PR              <- factor(df$PR, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$HER2            <- factor(df$HER2, levels = 0:1,
                               labels = c("Negative", "Positive"))
  df$surgery         <- factor(df$surgery, levels = 0:1,
                               labels = c("BCS", "Mastectomy"))
  df$chemo           <- factor(df$chemo, levels = 0:1,
                               labels = c("No", "Yes"))
  df$hormonal_therapy <- factor(df$hormonal_therapy, levels = 0:1,
                                labels = c("No", "Yes"))
  return(df)
}

datA <- convert_factors(datA)
datB <- convert_factors(datB)

str(datA)
```

## 2.2 Missing Data Check

```{r}
colSums(is.na(datA))
colSums(is.na(datB))
```

No missing values in either dataset — no imputation needed.

# 3 Exploratory Data Analysis (Cohort A)

## 3.1 Summary Statistics

```{r}
summary(datA[, -1])
```

## 3.2 Response Variable Distribution

```{r}
par(mfrow = c(1, 2))

hist(datA$log_PFS, breaks = 30,
     main = "Histogram of log(PFS)",
     xlab = "log(PFS)", col = "steelblue", border = "white")
abline(v = mean(datA$log_PFS), col = "red", lwd = 2, lty = 2)
legend("topright",
       legend = paste("Mean =", round(mean(datA$log_PFS), 3)),
       lty = 2, col = "red", lwd = 2, bty = "n")

qqnorm(datA$log_PFS, main = "Q-Q Plot of log(PFS)")
qqline(datA$log_PFS, col = "red", lwd = 2)
```

log(PFS) is approximately normal with mean `r round(mean(datA$log_PFS), 3)` and SD `r round(sd(datA$log_PFS), 3)`.

## 3.3 Continuous Predictors

```{r, fig.height=4}
cont_vars <- c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")

par(mfrow = c(1, 5), mar = c(4, 3, 2, 1))
for (v in cont_vars) {
  boxplot(datA[[v]], main = v, col = "steelblue", outline = TRUE)
}
```

```{r}
cont_df <- datA[, cont_vars]
data.frame(
  Variable = cont_vars,
  Mean     = round(sapply(cont_df, mean), 2),
  SD       = round(sapply(cont_df, sd), 2),
  Median   = round(sapply(cont_df, median), 2),
  Min      = round(sapply(cont_df, min), 2),
  Max      = round(sapply(cont_df, max), 2)
)
```

## 3.4 Correlation Matrix

```{r, fig.height=5}
cont_with_y <- c(cont_vars, "log_PFS")
cor_mat <- cor(datA[, cont_with_y])

corrplot(cor_mat,
         method  = "color",
         type    = "lower",
         addCoef.col = "black",
         number.cex  = 0.8,
         tl.col  = "black",
         tl.cex  = 0.9,
         title   = "Correlation Matrix (Continuous Variables)",
         mar     = c(0, 0, 2, 0))
```

Key correlations with `log_PFS`:

- **tumor_size** (r = `r round(cor_mat["tumor_size", "log_PFS"], 3)`) — moderate negative
- **lymph_nodes** (r = `r round(cor_mat["lymph_nodes", "log_PFS"], 3)`) — moderate negative
- **Ki67** (r = `r round(cor_mat["Ki67", "log_PFS"], 3)`) — moderate negative

Note: `tumor_size` and `lymph_nodes` are strongly correlated with each other (r = `r round(cor_mat["tumor_size", "lymph_nodes"], 3)`), suggesting potential multicollinearity.

## 3.5 Categorical Variable Distributions

```{r, fig.height=8}
cat_vars <- c("menopause", "race", "tumor_grade", "stage",
              "ER", "PR", "HER2",
              "surgery", "chemo", "hormonal_therapy")

par(mfrow = c(3, 4), mar = c(4, 3, 2, 1))
for (v in cat_vars) {
  barplot(table(datA[[v]]),
          main = v, col = "steelblue", border = "white",
          las = 2, cex.names = 0.8)
}
```

```{r}
lapply(cat_vars, function(v) {
  tbl <- table(datA[[v]])
  pct <- round(prop.table(tbl) * 100, 1)
  data.frame(Variable = v, Level = names(tbl), N = as.integer(tbl), Pct = paste0(pct, "%"))
}) |> do.call(rbind, args = _)
```

## 3.6 Outcome by Key Predictors

```{r, fig.height=8}
key_cats <- c("ER", "PR", "HER2", "tumor_grade", "stage", "hormonal_therapy")

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (v in key_cats) {
  boxplot(log_PFS ~ datA[[v]], data = datA,
          main = paste("log(PFS) by", v),
          xlab = v, ylab = "log(PFS)",
          col = "steelblue", border = "gray30")
}
```

## 3.7 Scatterplots: Continuous Predictors vs log(PFS)

```{r, fig.height=4}
par(mfrow = c(1, 5), mar = c(4, 4, 2, 1))
for (v in cont_vars) {
  plot(datA[[v]], datA$log_PFS,
       pch = 16, cex = 0.4, col = rgb(0, 0.3, 0.6, 0.3),
       xlab = v, ylab = "log(PFS)",
       main = paste(v, "vs log(PFS)"))
  abline(lm(datA$log_PFS ~ datA[[v]]), col = "red", lwd = 2)
}
```

---

*Next steps: model training, diagnostics, and external validation.*

# 4 Model Training

## 4.1 Prepare data for modeling
```{r}
# factor non-continuous variables *yw*
datA$race <- as.factor(datA$race)
datA$menopause <- as.factor(datA$menopause)
datA$tumor_grade <- as.factor(datA$tumor_grade)
datA$stage <- as.factor(datA$stage)

binary_vars <- c("ER", "PR", "HER2", "surgery", "chemo", "hormonal_therapy")
datA[binary_vars] <- lapply(datA[binary_vars], as.factor)

# creating train/test split
set.seed(1)
n = nrow(datA)
splitA = initial_split(datA, prop = 0.8)
trainA = training(splitA)
trainA = subset(trainA, select = -id)
testA = testing(splitA)
testA = subset(testA, select = -id)
ctrl = trainControl(method = "cv", number = 10)
```

## 4.2 Fitting the Linear Regression
```{r}
set.seed(1)
fit_lm = train(log_PFS ~ ., data = trainA, method = "lm", trControl = ctrl)
fit_lm
summary(fit_lm$finalModel)
pred_lm = predict(fit_lm, newdata = testA)
lm_result = postResample(pred = pred_lm, obs = testA$log_PFS)
```

## 4.3 Elastic Net
```{r}
set.seed(1)
fit_glmnet = train(log_PFS ~ ., data = trainA, method = "glmnet", trControl = ctrl, tuneLength = 10)
summary(fit_glmnet)
plot(fit_glmnet)

pred_glmnet = predict(fit_glmnet, newdata = testA)
glmnet_result = postResample(pred = pred_glmnet, obs = testA$log_PFS)
```

## 4.4 Interactive Elastic Net
```{r}
# *yw*
set.seed(1)
fit_glmnet_int <- train(log_PFS ~ .^2, data = trainA, method = "glmnet", trControl = ctrl, tuneLength = 10)
summary(fit_glmnet_int)
plot(fit_glmnet_int)

pred_glmnet_int <- predict(fit_glmnet_int, newdata = testA)
glmnet_int_result <- postResample(pred = pred_glmnet_int, obs = testA$log_PFS)
```

## 4.5 GAM Model
```{r}
# *yw*

set.seed(1)
fit_gam <- gam(
  log_PFS ~ s(age) + s(bmi) + s(tumor_size) + s(lymph_nodes) + s(Ki67) + 
            race + menopause + tumor_grade + stage + ER + PR + HER2 + 
            surgery + chemo + hormonal_therapy,
  data = trainA,
  method = "REML"
)
summary(fit_gam)

pred_gam <- predict(fit_gam, newdata = testA)
gam_result <- postResample(pred = pred_gam, obs = testA$log_PFS)
```


## 4.6 Model Comparison
```{r}
results <- data.frame(lm_result, glmnet_result, glmnet_int_result, gam_result)
results

# coef(fit_glmnet$finalModel, fit_glmnet$bestTune$lambda)

coefs <- coef(fit_glmnet_int$finalModel, fit_glmnet_int$bestTune$lambda)
coef_mat <- as.matrix(coefs)
active_coefs <- coef_mat[coef_mat[, 1] != 0, , drop = FALSE]
sorted_coefs <- active_coefs[order(-abs(active_coefs[, 1])), , drop = FALSE]

print(sorted_coefs)
```

## 4.7 Cross-Validation Summary & Tuning Parameters

```{r}
# *ms* CV performance summary across all models
cv_summary <- data.frame(
  Model = c("OLS", "Elastic Net", "Interactive Elastic Net", "GAM (test only)"),
  CV_RMSE = c(
    min(fit_lm$results$RMSE),
    min(fit_glmnet$results$RMSE),
    min(fit_glmnet_int$results$RMSE),
    NA
  ),
  CV_Rsquared = c(
    max(fit_lm$results$Rsquared),
    max(fit_glmnet$results$Rsquared),
    max(fit_glmnet_int$results$Rsquared),
    NA
  ),
  Test_RMSE = c(lm_result["RMSE"], glmnet_result["RMSE"],
                glmnet_int_result["RMSE"], gam_result["RMSE"]),
  Test_Rsquared = c(lm_result["Rsquared"], glmnet_result["Rsquared"],
                    glmnet_int_result["Rsquared"], gam_result["Rsquared"]),
  Test_MAE = c(lm_result["MAE"], glmnet_result["MAE"],
               glmnet_int_result["MAE"], gam_result["MAE"])
)
cv_summary[, -1] <- round(cv_summary[, -1], 4)
cv_summary
```

```{r}
# *ms* Best tuning parameters for elastic net models
best_glmnet <- fit_glmnet$bestTune
best_glmnet_int <- fit_glmnet_int$bestTune

data.frame(
  Model = c("Elastic Net", "Interactive Elastic Net"),
  alpha = c(best_glmnet$alpha, best_glmnet_int$alpha),
  lambda = round(c(best_glmnet$lambda, best_glmnet_int$lambda), 5)
)
```

## 4.8 Model Diagnostics (Interactive Elastic Net)

```{r}
# *ms* Residual diagnostics for the selected model
# Since caret's glmnet doesn't store residuals directly,
# we compute fitted values on training data
pred_train <- predict(fit_glmnet_int, newdata = trainA)
resid_train <- trainA$log_PFS - pred_train

par(mfrow = c(1, 3))

# Residuals vs Fitted
plot(pred_train, resid_train,
     pch = 16, cex = 0.5, col = rgb(0, 0.3, 0.6, 0.3),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lwd = 2, lty = 2)
lines(lowess(pred_train, resid_train), col = "blue", lwd = 2)

# Q-Q Plot of Residuals
qqnorm(resid_train, pch = 16, cex = 0.5,
       col = rgb(0, 0.3, 0.6, 0.5),
       main = "Normal Q-Q Plot of Residuals")
qqline(resid_train, col = "red", lwd = 2)

# Histogram of Residuals
hist(resid_train, breaks = 30,
     main = "Distribution of Residuals",
     xlab = "Residuals", col = "steelblue", border = "white")
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

Overall, the diagnostic review suggested that the model fit was reasonable for a predictive analysis. The residual plots were primarily used to check whether there were obvious patterns indicating systematic misspecification, such as strong nonlinearity or severe heteroscedasticity. The Q–Q plot and histogram were used to assess whether the residual distribution was approximately symmetric and centered around zero.

```{r}
# *ms* Shapiro-Wilk normality test on residuals (sample of 5000 if needed)
if (length(resid_train) > 5000) {
  set.seed(1)
  shapiro.test(sample(resid_train, 5000))
} else {
  shapiro.test(resid_train)
}
```

## 4.9 Result

Fitting the model with the interactive glmnet demonstrated superior predictive performance because it has the lowest RMSE and MAE among all tested models. We selected the interactive elastic-net as the final prediction model for Cohort A. The regularization penalty effectively managed the high-dimensional feature space, producing a parsimonious model by shrinking the vast majority of the non-informative main effects and potential interactions to exactly zero. The strongest standalone predictors that remained important included ER status, hormonal therapy, lymph node involvement, tumor grade, and Ki-67 index. Crucially, the model successfully isolated highly predictive interaction terms, capturing complex clinical synergies like ERPositive:hormonal_therapyYes and compounding risk factors such as HER2Positive:stageII and tumor_size:tumor_gradePoor.

## 4.10 Interpretation

Overall, the final model suggests that while aggressive tumor biology is associated with shorter progression-free survival, ER positivity and hormonal therapy act as strong baseline protective factors. Furthermore, the interaction terms reveal critical clinical finding: hormonal therapy provides a compounding, synergistic survival benefit specifically for ER-positive and Stage II patients, whereas the combination of large, poorly differentiated tumors drastically accelerates disease progression. However, overall predictive performance remains modest (test-set R-squared of 0.18). This indicates that while the inclusion of interaction terms successfully captures deeper biological dynamics and improves upon the base model, the available baseline variables still only explain a portion of the total variation in log_PFS.

For the next part, the elastic-net model from Cohort A would be carried forward for external validation on Cohort B.

# 5 Generalization

## 5.0 Cohort Comparison

```{r}
# *ms* Compare distributions between Cohort A and Cohort B
cohort_comp <- data.frame(
  Variable = c("n", "log_PFS mean (SD)", "Age mean (SD)", "BMI mean (SD)",
               "Tumor size mean (SD)", "Lymph nodes mean (SD)", "Ki67 mean (SD)",
               "ER positive (%)", "PR positive (%)", "HER2 positive (%)",
               "Chemo (%)", "Hormonal therapy (%)", "Mastectomy (%)"),
  Cohort_A = c(
    nrow(datA),
    paste0(round(mean(datA$log_PFS), 3), " (", round(sd(datA$log_PFS), 3), ")"),
    paste0(round(mean(datA$age), 1), " (", round(sd(datA$age), 1), ")"),
    paste0(round(mean(datA$bmi), 1), " (", round(sd(datA$bmi), 1), ")"),
    paste0(round(mean(datA$tumor_size), 2), " (", round(sd(datA$tumor_size), 2), ")"),
    paste0(round(mean(datA$lymph_nodes), 2), " (", round(sd(datA$lymph_nodes), 2), ")"),
    paste0(round(mean(datA$Ki67), 1), " (", round(sd(datA$Ki67), 1), ")"),
    round(mean(datA$ER == "Positive") * 100, 1),
    round(mean(datA$PR == "Positive") * 100, 1),
    round(mean(datA$HER2 == "Positive") * 100, 1),
    round(mean(datA$chemo == "Yes") * 100, 1),
    round(mean(datA$hormonal_therapy == "Yes") * 100, 1),
    round(mean(datA$surgery == "Mastectomy") * 100, 1)
  ),
  Cohort_B = c(
    nrow(datB),
    paste0(round(mean(datB$log_PFS), 3), " (", round(sd(datB$log_PFS), 3), ")"),
    paste0(round(mean(datB$age), 1), " (", round(sd(datB$age), 1), ")"),
    paste0(round(mean(datB$bmi), 1), " (", round(sd(datB$bmi), 1), ")"),
    paste0(round(mean(datB$tumor_size), 2), " (", round(sd(datB$tumor_size), 2), ")"),
    paste0(round(mean(datB$lymph_nodes), 2), " (", round(sd(datB$lymph_nodes), 2), ")"),
    paste0(round(mean(datB$Ki67), 1), " (", round(sd(datB$Ki67), 1), ")"),
    round(mean(datB$ER == "Positive") * 100, 1),
    round(mean(datB$PR == "Positive") * 100, 1),
    round(mean(datB$HER2 == "Positive") * 100, 1),
    round(mean(datB$chemo == "Yes") * 100, 1),
    round(mean(datB$hormonal_therapy == "Yes") * 100, 1),
    round(mean(datB$surgery == "Mastectomy") * 100, 1)
  )
)
cohort_comp
```

```{r}
# *ms* t-tests for continuous variables, chi-sq for categorical
cont_tests <- sapply(c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67", "log_PFS"), function(v) {
  round(t.test(datA[[v]], datB[[v]])$p.value, 4)
})
data.frame(Variable = names(cont_tests), p_value = cont_tests)
```

## 5.1 External Validation with Cohort B
```{r}
datB$race <- as.factor(datB$race)
datB$menopause <- as.factor(datB$menopause)
datB$tumor_grade <- as.factor(datB$tumor_grade)
datB$stage <- as.factor(datB$stage)

binary_vars <- c("ER", "PR", "HER2", "surgery", "chemo", "hormonal_therapy")
datB[binary_vars] <- lapply(datB[binary_vars], as.factor)

set.seed(1)
splitB <- initial_split(datB, prop = 0.8)
trainB <- subset(training(splitB), select = -id)
testB <- subset(testing(splitB), select = -id)

# predict the new 2000-2005 cohort using previous model
pred_b <- predict(fit_glmnet_int, newdata = testB)
external_validation_results <- postResample(pred = pred_b, obs = testB$log_PFS)

print(external_validation_results)
```

```{r}
# *ms* Predicted vs Actual plot for external validation
par(mfrow = c(1, 2))

# Model A on Cohort A test set
plot(testA$log_PFS, pred_glmnet_int,
     pch = 16, cex = 0.5, col = rgb(0, 0.3, 0.6, 0.4),
     xlab = "Observed log(PFS)", ylab = "Predicted log(PFS)",
     main = "Model A on Cohort A (Test)")
abline(0, 1, col = "red", lwd = 2, lty = 2)

# Model A on Cohort B test set
plot(testB$log_PFS, pred_b,
     pch = 16, cex = 0.5, col = rgb(0.6, 0.1, 0.1, 0.4),
     xlab = "Observed log(PFS)", ylab = "Predicted log(PFS)",
     main = "Model A on Cohort B (External)")
abline(0, 1, col = "red", lwd = 2, lty = 2)
```

The interactive Elastic Net model does not seem to be robust on cohort B as the RMSE increased from 0.74 to 0.87 and the MAE increased from 0.58 to 0.70. Since our is log-transformed PFS, a 0.12 increase in average error on the log scale is significant. The model trained on cohort A are missing the true survival times in cohort B by a wider margin. Considering factors like health care standards between the 1990s and the early 2000s, we should build a model to adapt the potential changes of predictors.

## 5.2 Fitting a new Interactive Elastic Net
```{r}
set.seed(1)
fit_glmnet_int_b <- train(log_PFS ~ .^2, data = trainB, method = "glmnet", trControl = ctrl, tuneLength = 10)

pred_b_new <- predict(fit_glmnet_int_b, newdata = testB)
new_model_results <- postResample(pred = pred_b_new, obs = testB$log_PFS)
print(new_model_results)

coefs_b <- coef(fit_glmnet_int_b$finalModel, fit_glmnet_int_b$bestTune$lambda)
coef_mat_b <- as.matrix(coefs_b)
active_coefs_b <- coef_mat_b[coef_mat_b[, 1] != 0, , drop = FALSE]
sorted_coefs_b <- active_coefs_b[order(-abs(active_coefs_b[, 1])), , drop = FALSE]

print(sorted_coefs_b)
```

```{r}
# *ms* CV summary and tuning for new Cohort B model
best_b <- fit_glmnet_int_b$bestTune
data.frame(
  Metric = c("Best alpha", "Best lambda", "CV RMSE", "CV R-squared",
             "Test RMSE", "Test R-squared", "Test MAE"),
  Value = round(c(
    best_b$alpha, best_b$lambda,
    min(fit_glmnet_int_b$results$RMSE),
    max(fit_glmnet_int_b$results$Rsquared),
    new_model_results["RMSE"], new_model_results["Rsquared"], new_model_results["MAE"]
  ), 4)
)
```

```{r}
# *ms* Diagnostics for new Cohort B model
pred_train_b <- predict(fit_glmnet_int_b, newdata = trainB)
resid_train_b <- trainB$log_PFS - pred_train_b

par(mfrow = c(1, 3))

plot(pred_train_b, resid_train_b,
     pch = 16, cex = 0.5, col = rgb(0, 0.3, 0.6, 0.3),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted (Cohort B)")
abline(h = 0, col = "red", lwd = 2, lty = 2)
lines(lowess(pred_train_b, resid_train_b), col = "blue", lwd = 2)

qqnorm(resid_train_b, pch = 16, cex = 0.5,
       col = rgb(0, 0.3, 0.6, 0.5),
       main = "Q-Q Plot (Cohort B)")
qqline(resid_train_b, col = "red", lwd = 2)

hist(resid_train_b, breaks = 30,
     main = "Residual Distribution (Cohort B)",
     xlab = "Residuals", col = "steelblue", border = "white")
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

```{r}
# *ms* Predicted vs Actual for new model
par(mfrow = c(1, 1))
plot(testB$log_PFS, pred_b_new,
     pch = 16, cex = 0.5, col = rgb(0, 0.3, 0.6, 0.4),
     xlab = "Observed log(PFS)", ylab = "Predicted log(PFS)",
     main = "New Model on Cohort B (Test)")
abline(0, 1, col = "red", lwd = 2, lty = 2)
```

There are several key changes in predictive terms:

- The intercept changed from 2.37 to 2.87, meaning the baseline PFS for reference patient improved from roughly 10.7 months to 17.6 months.

- ERPositive dropped from +0.196 to +0.107, while hormonal_therapyYes increased from 0.062 to 0.198. This flip proved that in 2000s, having a hormonal therapy is a more powerful driver of survival than simply having a less aggressive (ER+) tumor.

- The new model relies heavily on race-based interaction terms to make accurate predictions. i.e. raceAsian:stageII (-0.116) emerged as a risk factor, while raceHispanic:HER2Positive (+0.081) and raceHispanic:PRPositive (+0.026) emerged as protective. When care standard improves, unequal access to treatments between demographic groups become much stronger predictors of survival.

- The main effect for lymph_nodes (which was strongly negative in the 90s) shrinks to zero, meaning that finding a positive lymph node in early-stage might not be a death sentence anymore due to better medical treatment.

- In the old model, HER2Positive:stageII was a risk factor (-0.085), but it shrinks to zero in the new model. Now we see positive interactions like tumor_gradeModerate:HER2Positive (+0.024) and raceHispanic:HER2Positive (+0.081). This could be a result of the common use of Trastuzumab, a targeted therapy for HER2+ breast cancer.

